---
title: "Evolution and dynamics of case fatality rate in the COVID19 pandemia"
author: "J. J. Merelo"
date: "Sept 17th, 2020"
output:
  html_document: default
  pdf_document: default
bibliography: covid.bib
abstract: Different countries have different policies regarding testing of COVID-19 infections and reporting of deceases due to that infection. These policies show themselves in the evolution of the case fatality ratio, or relation between total number of infected people and reported deceases. Its dynamics and statistical features will allow us to assess these policies mainly regarding under-testing or under-reporting of fatalities.
urlcolor: blue
---

```{r setup, include=FALSE}
library(tidyr)
library(ggplot2)
library(ggthemes)
library(dplyr)

base <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_"
world.confirmed <- read.csv(paste0(base,"confirmed_global.csv"), sep=',',head=T)
world.confirmed <- gather( world.confirmed, Date, Cases, X1.22.20:X9.15.20)
world.deaths <- read.csv(paste0(base,"deaths_global.csv"), sep=',',head=T)
world.deaths <- gather( world.deaths, Date, Deaths, X1.22.20:X9.15.20)
world.data <- merge(world.confirmed,world.deaths,by=c("Province.State","Country.Region","Lat", "Long", "Date"))
world.data$Date <- as.Date(world.data$Date,"X%m.%d.%y")
world.data <- world.data %>% group_by(Province.State,Country.Region,Date) %>% arrange(Province.State,Country.Region, as.Date(Date))
world.data <- world.data %>% group_by(Province.State,Country.Region) %>% mutate( New.Cases= Cases-lag(Cases, default=0), New.Deaths= Deaths -lag(Deaths, default=0))
save(world.data,file="covid19-world.Rda")
write.csv(world.data,'covid19-world.csv')
```

# Introduction

One of the essential weapons to fight pandemic is information. Knowing in a precise way how different countries have dealt with COVID-19 to ascertain how it has impacted case and mortality allows new countries and territories to deal with it in the most adequate way. 

Trying to ascertain the real number of infected people,  deceases stemming from the infection and their evolution is one of those things, and countries have reacted differently to this challenge. From testing only those with symptoms and that self-select and call up health services, to testing massively, as it has been done in South Korea [@SHIM2020], Germany, or in Japan, at least with people repatriated from affected areas [@nishiura2020rate]. This might be the reason why these countries report a lower Case Fatality Ratio (CFR) than in other cases.

Let's first check what's the reported case fatality ratio for different regions in the world, those that have at least 1000 cases. Max and min CFR are computed over the rows in which there were already 1000 cases; `last.CFR` is the last case fatality ratio that has been reported. Territories are sorted over the last CFR reported.

```{r cfr,echo=FALSE,message=FALSE}
world.data$CFR <- world.data$Deaths / world.data$Cases
world.data.filtered.1k <- world.data[ world.data$Cases > 500000, ]
world.data.filtered.1k %>% group_by(Country.Region,Province.State) %>% summarize( max.CFR = max(CFR), min.CFR = min(CFR), last.CFR = last(CFR))  %>% arrange( - last.CFR) %>% rmarkdown::paged_table()
```

The countries with the lowest CFR have peaks in the area of 0.3%, with some cases even below that; China and Germany have very low last reported Cf Rs. The two mentioned countries, South Korea and Germany, have a current CFR in the 0.2-0.8% area. This contrasts with the US, which has the highest current CFR at 5%, similar to Italy, the UK and Netherlands.

The evolution of CFRs is presented in this chart

```{r cfrevol,echo=FALSE,message=FALSE}
library(viridis)
world.data.filtered.1k$Territory <- paste(world.data.filtered.1k$Province.State, world.data.filtered.1k$Country.Region)
ggplot(world.data.filtered.1k, aes(x=Date,y=CFR,color=Territory,group=Territory))+geom_line()+scale_color_viridis_d()+theme_tufte()+theme(legend.position = "bottom")+labs(title = "Evolution of case fatality ratio",subtitle = "Countries and territories with more than 1k cases")
```

This chart, that shows the evolution of the CFR once 1000 cases have been reached in a territory, shows roughly two groups. In one, the CFR remains roughly constant after initial growth; the other shows unlimited growth after the beginning. To exemplify these behaviors, let's show Switzerland vs. Italy:

```{r de.it,echo=FALSE,message=FALSE}
just.de.it <- world.data.filtered.1k[ world.data.filtered.1k$Country.Region == "Switzerland" | world.data.filtered.1k$Country.Region == "Italy",]
ggplot(just.de.it, aes(x=Date,y=CFR,color=Territory,group=Territory))+geom_line()+theme_tufte()+theme(legend.position = "bottom")
```

While the CFR remains low, and more or less constant, in Switzerland, the one in Italy grows as just the cases that enter hospital are tested for coronavirus, yielding an ever-increasing CFR, which can't simply be true since it's unrelated to the individual fatality rates reported in the literature following single cases, which should be around 0.5% [@hsunovel]


# Conclusions

We can draw two main conclusions from this visualization of case fatality rates. Successful policies show an approximately flat line, with case-to-decease ration remaining approximately constant, which indicates that treatment facilities have not been overcome and every case is receiving the treatment it needs.

Non-constant, and growing, CFRs clearly indicate a failure in policies; an ever growing number of deaths (related to the total number of reported cases) might indicate either lack of reporting of cases, overflowing of treatment facilities, or both. The main conclusion is that extensive testing, or lacking tests, some other kind of self-reporting of cases or nowcasting the number of existing cases is essential for countries and territories to properly deal with the pandemic.


## Acknowledgements

This file has been generated from data published by <a href="https://github.com/CSSEGISandData/COVID-19" target="_new">JHU CSSE</a>. It's data-driven and it can be re-generated from the script in  <a href="https://github.com/JJ/covid-reports" target="_new">this repository</a>. This report uses RMarkdown and it can be downloaded from <a href="https://github.com/JJ/covid-reports/blob/master/cfr-evolution.Rmd" target="_new">here</a>.

## References
