---
title: "Estimating the real number of cases of COVID19 in Spain and elsewhere"
author: "J. J. Merelo"
date: "21 de marzo de 2020"
bibliography: covid.bib
urlcolor: blue
output:
  html_document: default
  pdf_document: default
abstract:
  In a pandemic such as this one, it's impossible to ascertain the real number of cases except by testing the whole population; this leads to the problem of different death rates for different countries, but mainly to the more poignant problem of allocating resources for urgent and grave cases or assessing the local peak of the epidemic. In this report we will try to find out an estimate of the death rate by looking at the countries that have made a more extensive testing, like Germany or South Korea. We will first try to estimate the time from onset to outcome, and from that, we will try and give an estimate of cases (and possible deaths) in other countries where testing has not been so extensive, like Spain.
---

```{r setup, include=FALSE}
library(tidyr)

base <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_19-covid-"
world.confirmed <- read.csv(paste0(base,"Confirmed.csv"), sep=',',head=T)
world.confirmed <- gather( world.confirmed, Date, Cases, X1.22.20:X3.20.20)
world.deaths <- read.csv(paste0(base,"Deaths.csv"), sep=',',head=T)
world.deaths <- gather( world.deaths, Date, Deaths, X1.22.20:X3.20.20)
world.data <- merge(world.confirmed,world.deaths,by=c("Province.State","Country.Region","Lat", "Long", "Date"))
world.data$Date <- as.Date(world.data$Date,"X%m.%d.%y")
save(world.data,file="covid19-world.Rda")
write.csv(world.data,'covid19-world.csv')
```

# Introduction

Transparency in the management of a critical situation such as the one we are living with coronavirus is essential. Not only for the peace of mind of the population, but also for being able to take informed decisions on the allocation of resources for those affected with the pandemia.

Knowing the real number of infected people and their evolution is one of those things, and countries have reacted differently to this challente. From testing only those with symtomps and that self-select and call up health services, to testing massively, as it has been done in South Korea [@SHIM2020], Germany, or in Japan, at least with people repatriated from affected areas [@nishiura2020rate]. This might be the reason why these countries report a lower Case Fatality Ratio (CFR) than in other cases.

Let's first check what's the reported case fatality ratio for different regions in the world, those that have at least 1000 cases. Max and min CFR are computed over the rows in which there were already 1000 cases.

```{r cfr,echo=FALSE,message=false}
library(dplyr)
world.data$CFR <- world.data$Deaths / world.data$Cases
world.data.filtered.1k <- world.data[ world.data$Cases > 1000, ]
world.data.filtered.1k %>% group_by(Country.Region) %>% summarize( max.CFR = max(CFR), min.CFR = min(CFR), last.CFR = last(CFR))  %>% arrange( - last.CFR)
```

The countries with the lowest CFR is in the area of 0.3%, with some cases even below that. 

## Acknowledgements

This file has been generatd from data published by  [Datadista](https://github.com/datadista/datasets) as well as [JHU CSSE](https://github.com/CSSEGISandData/COVID-19). It's data-driven and it can be re-generated from the script in  [this repository](https://github.com/JJ/covid-reports).
