---
title: "Estimating the real number of cases of COVID19 in Spain and elsewhere"
author: "J. J. Merelo"
date: "21 de marzo de 2020"
bibliography: covid.bib
urlcolor: blue
output:
  html_document: default
  pdf_document: default
abstract:
  In a pandemic such as this one, it's impossible to ascertain the real number of cases except by testing the whole population; this leads to the problem of different death rates for different countries, but mainly to the more poignant problem of allocating resources for urgent and grave cases or assessing the local peak of the epidemic. In this report we will try to find out an estimate of the death rate by looking at the countries that have made a more extensive testing, like Germany or South Korea. We will first try to estimate the time from onset to outcome, and from that, we will try and give an estimate of cases (and possible deaths) in other countries where testing has not been so extensive, like Spain.
---

```{r setup, include=FALSE}
library(tidyr)

base <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_19-covid-"
world.confirmed <- read.csv(paste0(base,"Confirmed.csv"), sep=',',head=T)
world.confirmed <- gather( world.confirmed, Date, Cases, X1.22.20:X3.20.20)
world.deaths <- read.csv(paste0(base,"Deaths.csv"), sep=',',head=T)
world.deaths <- gather( world.deaths, Date, Deaths, X1.22.20:X3.20.20)
world.data <- merge(world.confirmed,world.deaths,by=c("Province.State","Country.Region","Lat", "Long", "Date"))
world.data$Date <- as.Date(world.data$Date,"X%m.%d.%y")
world.data$New.Cases <-  c( NA, diff(world.data$Cases))
world.data$New.Deaths <-  c( NA, diff(world.data$Deaths))
save(world.data,file="covid19-world.Rda")
write.csv(world.data,'covid19-world.csv')
```

# Introduction

Transparency in the management of a critical situation such as the one we are living with coronavirus is essential. Not only for the peace of mind of the population, but also for being able to take informed decisions on the allocation of resources for those affected with the pandemia.

Knowing the real number of infected people and their evolution is one of those things, and countries have reacted differently to this challente. From testing only those with symtomps and that self-select and call up health services, to testing massively, as it has been done in South Korea [@SHIM2020], Germany, or in Japan, at least with people repatriated from affected areas [@nishiura2020rate]. This might be the reason why these countries report a lower Case Fatality Ratio (CFR) than in other cases.

Let's first check what's the reported case fatality ratio for different regions in the world, those that have at least 1000 cases. Max and min CFR are computed over the rows in which there were already 1000 cases.

```{r cfr,echo=FALSE,message=FALSE}
library(dplyr)
world.data$CFR <- world.data$Deaths / world.data$Cases
world.data.filtered.1k <- world.data[ world.data$Cases > 1000, ]
world.data.filtered.1k %>% group_by(Country.Region) %>% summarize( max.CFR = max(CFR), min.CFR = min(CFR), last.CFR = last(CFR))  %>% arrange( - last.CFR)
```

The countries with the lowest CFR have peaks in the area of 0.3%, with some cases even below that; China and Germany have very low last reported CFRs. The two mentioned countries, South Korea and Germany, have a current CFR in the 0.2-0.8% area. This constrasts with the US, which has the highest current CFR at 5%, similar to Italy, the UK and Netherlands.

But the case fatality ratio does not give the whole picture. Some might have been tested one day before, other even post-mortem. Another quantity, the infected fatality ratio, will give a more accurate scenario of what's happening. But in absence of individualized data, we need to deduce that from published data, by calculating correlations between cases and deaths. We'll do this next.

# Correlation between cases and deaths.

What we need to know is, approximately, what's the expected time it transcurs from infection to the final outcome. We'll zero in on Germany and South Korea for this. Let's plot correlation first for South Korea

```{r acf.ko,echo=FALSE,message=FALSE}
KO.data <- world.data[world.data$Country.Region=="Korea, South",]
ccf(KO.data$New.Cases, KO.data$New.Deaths,lag.max = 28)

```

There's negative correlation 12 and 2 days before, as well as positive same-day and -3 days. That is, lower than average cases will lead to higher-than-average deaths 12 days later. 

Let's do the same for Germany:
```{r acf.de,echo=FALSE,message=FALSE}
DE.data <- world.data[world.data$Country.Region=="Germany",]
ccf(KO.data$New.Cases, DE.data$New.Deaths,lag.max = 28)

```

There is a very strong positive correlation with a lag of 10 days, as well as a very strong negative correlation with a lag of 12 days and 9 days; same-day correlation is also strong, but not as high as in the case of Korea.

## Acknowledgements

This file has been generatd from data published by  [Datadista](https://github.com/datadista/datasets) as well as [JHU CSSE](https://github.com/CSSEGISandData/COVID-19). It's data-driven and it can be re-generated from the script in  [this repository](https://github.com/JJ/covid-reports).
